Can Auxiliary Indicators Improve COVID-19 Forecasts and Hotspot Prediction?

### 2021-06-12 forecast/paper/paper.tex
Edits almost thru methods B

\dropcap{T}racking and forecasting indicators from public health reporting1
streams---such as confirmed cases and deaths in the COVID-19 pandemic---is
crucial for understanding disease spread, formulating public policy responses,
and anticipating future public health resource needs.  In a companion paper, we
describe our research group's (Delphi's) efforts in curating and maintaining a 
database of real-time indicators that track COVID-19 activity and other relevant
phenomena. The signals (a term we use synonomously with ``indicators'') in this
database are accessible through the COVIDcast API \cite{CovidcastAPI}, with
associated R \cite{CovidcastR} and Python \cite{CovidcastPy} packages for
convenient data fetching and processing tools. In the current paper, we aim to
quantify the utility provided by a core set of these indicators for two
fundamental prediction tasks: probabilistic forecasting of COVID-19 case 
rates and prediction of future COVID-19 hotspots (where we define a hotspot by a
relative increase in COVID-19 cases, to be spelled out precisely later).

At the outset, we should be clear that our intent in this paper is \emph{not} to
provide an authoritative take on cutting-edge COVID-19 forecasting methods.
Instead, our purpose here is to provide a rigorous, quantitative assessment of
the utility that several auxiliary indicators---such as those derived from
internet surveys or medical insurance claims---provide in tasks that involve
predicting future trends in confirmed COVID-19 cases. To assess such utility in
as simple terms as possible, we center our study in the framework of a basic
autoregressive model (in which COVID cases in the near future are predicted from  
a linear combination of COVID cases in the near past), and ask whether the 
inclusion of an auxiliary indicator as an exogenous feature in such a model
improves predictive power. While forecasting carries a rich literature that
offers a wide range of techniques, see e.g., \cite{Hyndman:2018}, we purposely
constrain ourselves to very simple models, avoiding common enhancements such as
order selection, correction of outliers/anomalies in the data, and inclusion of
regularization or nonlinearities. That said, analyses of forecasts submitted to
the COVID-19 Forecast Hub \cite{ForecastHub} by a large community of modelers
have shown that simple, robust models have consistently been among the
best-performing over the pandemic \cite{Cramer:2021}, including time series
models similar to those we consider in what follows.

In our companion paper, we analyze correlations between various indicators and   
COVID case rates. These correlations are natural summaries of the
contemporaneous association between an indicator and COVID cases, but they fall
short of delivering a satisfactory answer to the question that motivates the
current article: is the information contained in an indicator demonstrably
useful for the prediction tasks we care about? For such a question, lagged
correlations (e.g., measuring the correlation between an indicator and COVID
case rates several days in the future) move us in the right direction, but still
do not move us all the way there. The question about \emph{utility for
  prediction} is focused on a much higher standard than simply asking 
about correlations; to be useful in forecast or hotspot models, an indicator
must provide relevant information that is not otherwise contained in past values
of the case rate series itself. We will assess this in the most direct way
possible: by inpsecting the difference in predictive performance of simple
autoregressive models trained with and without access to past values of a
particular indicator.

There is another, more subtle issue in evaluating predictive utility that 
deserves explicit mention, as it will play a key role in our analysis.
Signals computed from surveillance streams will often be subject to  
latency and/or revision. For example, a signal based on aggregated medical
insurance claims may be available after just a few days, but it can then be
substantially revised over the next several weeks as additional claims are
submitted and/or processed late. Correlations between such a signal and case
rates calculated ``after the fact'' (computed retrospectively, using the
finalized values of this signal) will not deliver an honest answer to the    
question of whether this signal would have been useful in real time. Instead,
we build predictive models using only the data that would have been available
\emph{as of} the prediction date, and compare the ensuing predictions in terms
of accuracy. This leverages our group's {\tt evalcast} R package, which plugs
into the COVIDcast API's data versioning system, and facilitates honest
backtesting of forecasters.

@@

The indicators we focus on provide information not generally available from
standard public health reporting. Among the many available in Delphi's COVIDcast
API, we consider the following five auxiliary indicators: 
\begin{itemize}
\item Change Healthcare COVID-like illness (CHNG-CLI): The percentage of
  outpatient visits that are primarily about COVID-related symptoms, based on
  de-identified Change Healthcare claims data.
\item Change Healthcare COVID (CHNG-COVID): The percentage of outpatient visits
  with confirmed COVID-19, based on the same claims data.
\item COVID Tracking and Impact Survey CLI-in-community (CTIS-CLI-in-community): 
  The estimated percentage of the population who knows someone who is sick,
  based on Delphi's surveys of Facebook users.
\item Doctor Visits COVID-like illness (DV-CLI): The same as CHNG-CLI, but
  computed based on de-identified medical insurance claims from other health
  systems partners.  
\item Google Search Trends for anosmia and ageusia (Google-AA): A measure of 
  COVID-19 related Google search volume for queries related to anosmia (loss of
  smell), ageusia (loss of taste), or both.  
\end{itemize}
In short, we choose these indicators because, conceptually speaking, they
measure aspects of an individual's disease progression that would plausibly
precede the occurence of (at worst, co-occur  with) the report of a positive
COVID-19 test, through standard public health reporting streams.

@@

As for geographic resolution, we consider the prediction of COVID-19 case rates
and hotspots (to be precisely defined shortly) aggregated at the level of an
individual hospital referral region (HRR). HRRs correspond to groups of counties
in the United States within the same hospital referral system. The Dartmouth
Atlas of Healthcare Policy~\cite{DartmouthHRR}, defines these 306 regions based
on a number of characteristics. They are contiguous regions containing at least
one city where major procedures (cardiovascular or neurological) are
performed. Each HRR has a minimum population of about 120,000, and the majority
of hospital services for that population are performed by hospitals within the
region. Some HRRs are quite large (such as the one containing Los Angeles, which
has about 10 million people), but generally HRRs are much more homogenous in
size than the (approximately) 3200 counties are, and offer a nice middle ground
in between counties and states.

HRRs, by their definition, would be most relevant for hospitalization
forecasting.  We have chosen to focus on forecasting cases at this resolution 
because the indicators that we consider should intuitively be more useful in
latter task.  Predicting case rates and case hotspots at the HRR level is still
a reasonable goal, especially predicted case information could then be fed
downstream into hospitalization forecasting models.

@@

% RJT: We need to fill this out.  I'm convinced we need to promote "as of"
% training as our main point of study.  We need to explain why this is so
% important.  And we can and should present results with finalized data in the
% appendix.  I guess we can wait until we get the experiments in to fill out
% this section, because it might influence the prose we use here, or we could
% also start writing this right away. And we can refer to the companion
% (indicators) paper, which also describes the revision tracking feature in the
% API.  @DM/JB: Your call, whether to start writing this or wait.
% DJM: drafted

In this manuscript, all models are estimated using only ``vintage training
data''. This means that for a given forecast date, say November 1st 2020, we
estimate our models using only the data that would have been available to us on
November 1st. Essentially, the training data
is rewound appropriately to dates in the past. It is likely, for example, that
no data would actually exist 
for the forecast date on the forecast date because there is some latency between
the time signals are reported and the dates for which they are reported.
Similarly, any future revisions made to the indicators or case numbers would
not have been present on the forecast date, and so would not be available to the
forecaster. Figure~\ref{fig:vintage} shows three signals for the Los Angeles HRR
with a hypothetical forecast date of November 1. On November 1, no signals are
available for October 31. The cases data is one day latent while the Facebook
survey is 2 days latent. Doctor's visits are 3 days latent while also exhibiting
revisions.  
By training our models in this way, the evaluations are more realistic in the sense
that we are accurately mimicking the difficulty faced by real-time epidemic
forecasters. The COVIDcast API maintains all versions of the data, thereby allowing
these sorts of synthetic forecasting exercises. For more details about data
versioning, see the companion paper.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{fig/vintage.pdf}
  \caption{Changes in availability and revisions for three signals in the Los
    Angeles HRR. The vertical black line indicates a hypothetical forecast date
    of November 1 while the colored lines get progressively darker as we collect
  later and later vintages of data.}
  \label{fig:vintage}
\end{figure}

@@

% RJT: Just a quick few quick comments for now on the text below.  
% Also, shouldn't we
% point out it's relationship to quantile score, and give it's equivalent form
% there?
% DJM: done.

@@

>>>>

\dropcap{T}racking and forecasting indicators from public health reporting
streams---such as confirmed cases and deaths in the COVID-19 pandemic---is
crucial for understanding disease spread, formulating public policy responses,
and anticipating future public health resource needs.  In a companion paper, we
describe our research group's (Delphi's) efforts in curating and maintaining a 
database of real-time indicators that track COVID-19 activity and other relevant
phenomena. The signals (a term we use synonomously with ``indicators'') in this
database are accessible through the COVIDcast API \cite{CovidcastAPI}, with
associated R \cite{CovidcastR} and Python \cite{CovidcastPy} packages for
convenient data fetching and processing tools. In the current paper, we aim to
quantify the utility provided by a core set of these indicators for two
fundamental prediction tasks: probabilistic forecasting of COVID-19 case 
rates and prediction of future COVID-19 case hotspots (defined by the event that
a relative increase in COVID-19 cases exceeds a certain threshold).

At the outset, we should be clear that our intent in this paper is \textit{not}
to provide an authoritative take on cutting-edge COVID-19 forecasting methods.
Instead, our purpose here is to provide a rigorous, quantitative assessment of
the utility that several auxiliary indicators---such as those derived from
internet surveys or medical insurance claims---provide in tasks that involve
predicting future trends in confirmed COVID-19 cases. To assess such utility in
as simple terms as possible, we center our study in the framework of a basic
autoregressive model (in which COVID cases in the near future are predicted from  
a linear combination of COVID cases in the near past), and ask whether the 
inclusion of an auxiliary indicator as an exogenous feature in such a model
improves predictive power. While forecasting carries a rich literature that
offers a wide range of techniques, see e.g., \cite{Hyndman:2018}, we purposely
constrain ourselves to very simple models, avoiding common enhancements such as
order selection, correction of outliers/anomalies in the data, and inclusion of
regularization or nonlinearities. That said, analyses of forecasts submitted to
the COVID-19 Forecast Hub \cite{ForecastHub} by a large community of modelers
have shown that simple, robust models have consistently been among the
best-performing over the pandemic \cite{Cramer:2021}, including time series
models similar to those we consider in what follows.

In our companion paper, we analyze correlations between various indicators and   
COVID case rates. These correlations are natural summaries of the
contemporaneous association between an indicator and COVID cases, but they fall 
short of delivering a satisfactory answer to the question that motivates the
current article: is the information contained in an indicator demonstrably
useful for the prediction tasks we care about? For such a question, lagged
correlations (e.g., measuring the correlation between an indicator and COVID
case rates several days in the future) move us in the right direction, but still
do not move us all the way there. The question about \textit{utility for 
  prediction} is focused on a much higher standard than simply asking 
about correlations; to be useful in forecast or hotspot models, an indicator
must provide relevant information that is not otherwise contained in past values
of the case rate series itself. We will assess this in the most direct way
possible: by inpsecting the difference in predictive performance of simple
autoregressive models trained with and without access to past values of a
particular indicator.

There is another, more subtle issue in evaluating predictive utility that 
deserves explicit mention, as it will play a key role in our analysis.
Signals computed from surveillance streams will often be subject to  
latency and/or revision. For example, a signal based on aggregated medical
insurance claims may be available after just a few days, but it can then be
substantially revised over the next several weeks as additional claims are
submitted and/or processed late. Correlations between such a signal and case
rates calculated ``after the fact'' (computed retrospectively, using the
finalized values of this signal) will not deliver an honest answer to the    
question of whether this signal would have been useful in real time. Instead,
we build predictive models using only the data that would have been available
\textit{as of} the prediction date, and compare the ensuing predictions in terms 
of accuracy. To do so, we leverage Delphi's \texttt{evalcast} R package 
\cite{EvalcastR}, which plugs into the COVIDcast API's data versioning system,  
and facilitates honest backtesting.

@@

We consider prediction of future COVID-19 case rates or case hotspots (to be
defined precisely shortly).  By case rate, we mean the case count per 100,000
people (the standard in epidemiology).  We use reported case data aggregated by 
JHU CSSE \cite{Dong:2020}, which is accessible through the COVIDcast API, like
the auxiliary indicators that we use to supplement the basic purely
autoregressive models.

The indicators we focus on provide information not generally available from
standard public health reporting. Among the many indicators in the API, we
consider the following five:
\begin{itemize}
\item Change Healthcare COVID-like illness (CHNG-CLI): The percentage of
  outpatient visits that are primarily about COVID-related symptoms, based on
  de-identified Change Healthcare claims data.
\item Change Healthcare COVID (CHNG-COVID): The percentage of outpatient visits
  with confirmed COVID-19, based on the same claims data.
\item COVID Tracking and Impact Survey CLI-in-community (CTIS-CLI-in-community): 
  The estimated percentage of the population who knows someone who is sick,
  based on Delphi's surveys of Facebook users.
\item Doctor Visits COVID-like illness (DV-CLI): The same as CHNG-CLI, but
  computed based on de-identified medical insurance claims from other health
  systems partners.  
\item Google Search Trends for anosmia and ageusia (Google-AA): A measure of 
  COVID-19 related Google search volume for queries related to anosmia (loss of
  smell), ageusia (loss of taste), or both.  
\end{itemize}
In short, we choose these indicators because, conceptually speaking, they
measure aspects of an individual's disease progression that would plausibly
precede the occurence of (at worst, co-occur  with) the report of a positive
COVID-19 test, through standard public health reporting streams.

@@

As for geographic resolution, we consider the prediction of COVID-19 case rates
and hotspots aggregated at the level of an individual \textit{hospital referral
  region} (HRR). HRRs correspond to groups of counties in the United States
within the same hospital referral system. The Dartmouth Atlas of Healthcare 
Policy~\cite{DartmouthHRR}, defines these 306 regions based on a number of
characteristics. They are contiguous regions containing at least one city where
major procedures (cardiovascular or neurological) are performed.  Most hospital
services for the population in an HRR are performed by hospitals within the
region.  Each HRR has a minimum population of about 120,000; some HRRs are quite
large (such as the one containing Los Angeles, which has about 10 million
people), but generally HRRs are much more homogenous in size than the
(approximately) 3200 counties are, and they serve as a nice middle ground in 
between counties and states.

HRRs, by their definition, would be most relevant for forecasting hospital
admissions.  We have chosen to focus on cases (forecasting and predicting
hotspots) at this resolution since the indicators considered should be more
useful in predicting case versus hospitalization activity, the latter being
intuitively more closely connected (in time) to the events that are measured by
the given five indicators. Predicting case rates (and hotspots) at the HRR level
is still a reasonable goal in its own right; and moreover, it could be used to
feed predicted case information into downstream hospitalization models.

@@

In this paper, all models are fit using only ``vintage'' training data. This
means that for a given prediction date, say September 1, 2020, we train models
using only the data that would have been available to us \textit{as of}
September 1.  Essentially, we ``rewind'' the clock to September 1, and query the
COVIDcast API to get the latest data that it would have had available at
that point in time.  This is made possible by the COVIDcast API's comprehensive
data versioning system (described in more detail in our companion paper).  We
also use the \texttt{evalcast} R package \cite{EvalcastR}, which streamlines the
process of training arbitrary prediction models over a sequence of prediction
dates, by automatically constructing the (properly-versioned) sequence of
vintage training data sets.

Vintage training data means different things, in practice, for different
signals. The three signals based on medical claims, CHNG-CLI, CHNG-COVID, and
DV-CLI, are most often 3-5 days latent, and subject to a considerable degree of
revision or ``backfill'' after their initial publication date.  The survey-based
signal, CTIS-CLI-in-community, is 2 days latent, and rarely undergoes any 
revision at all.  The target itself, reported case rates (or a binary hotspot
variable, defined based on reported case rates), is 1 day latent, and exhibits
frequent, unpredictable revisions (sometimes clear anomalies) after initial
publication.  Compared to the revision pattern in the medical claims signals,
which are appears to be much more systematic, the revision pattern in the case
rates signal can be highly erratic, resulting in big spikes as reporting
backlogs are cleared, changes in case definitions are made, other corrections
are applied, etc.

TODO What is ground truth??
TODO Better example of backfill in DV-CLI?

@@

@@

Code to reproduce all examples (which uses the
\texttt{evalcast} R package and fetches data directly from the API) can be
found at \url{TODO}.
